The Economics of Wildfire Aviation

### 2023-09-05 manuscript/manuscript.qmd
rearranged intro

Over the past 10 years, the federal government has spent an average of \$2.3 billion per year on wildfire suppression [@suppress]. Approximately 21% (\$493 million) of that is spent on aviation resources [@stonesifer2021]. These costs involve hourly flight costs to operate the aircraft, the cost of retardant, and the cost to keep aircraft available and maintain it. Given the high cost of aviation, it is critical to determine whether the benefits of their application exceed the costs. Otherwise, resources could be better invested in other firefighting resources such as wildland firefighting engines, dozers, and personnel.

For the past decade, federal firefighting agencies have contracted with private companies that own and maintain firefighting aircraft. The structure of these contracts and the procurement mechanism influence how aviation resources are used and the costs borne by firefighting agencies.

@@

The objective of this paper is to describe and analyze two economic aspects of aviation used for wildfire response. First, we discuss the lack of objective measurement of the benefits of aviation that is necessary to subject their use to benefit cost analysis. Second, we apply economic principles to the procurement of aviation resources and discuss how and where economic efficiency can be gained. While this paper will not analyze data or come to clear conclusions about the use of aviation, it will frame the economic problems and attempt to provide a path forward for how to solve them.

@@

Wildfire response is a coordinated effort between federal, state, and local land management agencies. Aviation resources have been used in wildfire response since WWII [@pyne1982], and represent an important yet expensive firefighting resource. Helicopters and fixed-wing aircraft are used to drop water and retardant on or in advance of the fire. Aircraft are also used to gather information about the extent and intensity of the fire, and to survey damage. While aviation can provide valuable services, they are one of the most expensive resources on an active fire.

@@

# Measuring the benefits of aviation

@@

# Procurement

@@

# Discussion

@@

# Maybe material

>>>>

Over the past 10 years, the federal government has spent an average of \$2.3 billion per year on wildfire suppression [@suppress]. These costs are expected to rise as climate change drives more intense fires over longer seasons [@burke2021; @abatzoglou2016]. Approximately 21% (\$493 million) of that is spent on aviation resources [@stonesifer2021]. These costs involve hourly flight costs to operate the aircraft, the cost of retardant, and the cost to keep aircraft available and maintain it. Given the high cost of aviation, it is critical to determine whether the benefits of their application exceed the costs. Otherwise, resources could be better invested in other firefighting resources such as wildland firefighting engines, dozers, and personnel.

The procurement of aviation resources influences federal expenditure on firefighting aviation. For the past decade, federal firefighting agencies have contracted with private companies that own and maintain firefighting aircraft. Currently, federal agencies either lease aircraft for much of the season or sign short term contracts for a higher daily cost. The structure and prices of these contracts and the procurement mechanism influence how aviation resources are used and the costs borne by firefighting agencies.

@@

The objective of this paper is to describe and analyze two economic aspects of aviation used for wildfire response. First, we discuss how the benefits of aviation are not well measured or tracked, making it difficulty to analyze the benefits relative to the costs. Second, we apply economic principles to the procurement of aviation resources and discuss how and where economic efficiency can be gained. While this paper will not come to clear conclusions about the use of aviation, it will frame the economic problems and attempt to provide a path forward for how to solve them.

## Background

@@

Wildfire response is a coordinated effort between federal, state, and local land management agencies. Aviation resources have been used in wildfire response since WWII [@pyne1982] when military aircraft were retrofit with equipment that would allow them to carry and drop water. Over the years, government agencies have acquired, retrofit, and maintained a fleet of airplanes and helicopters. These aircraft have been used to drop water and retardant on or in advance of the fire, deploy smoke jumpers, to gather information about the extent and intensity of the fire, and to survey damage. While aviation can provide valuable services, they are one of the most expensive resources on an active fire.

@@

## Measuring the benefits of aviation

@@

## Procurement

@@

## Discussion

@@

### Maybe material
